Tight Generalization Bounds for Hard Parameter Sharing in High-dimensional Linear Regression

### 2020-10-11 arxiv/diff_cov.tex
sec A

\begin{example}[Covariate shift]\label{ex_covshift}
%So far we have considered the isotropic model where $\Sigma_1 = \Sigma_2$.
%This setting is relevant for settings where different tasks share the same input features such as multi-class image classification.
%In general, the covariance matrices of the two tasks may be different such as in text classification.
Our second example studies the impact of varying covariate shifts on variance in the left hand side of equation \eqref{lem_cov_shift_eq}:
	\[ \bigtr{\Sigma^{(2)} \hat{\Sigma}^{-1}} \rightarrow \frac{1}{n_1 + n_2} \bigtr{\Sigma^{(2)} (a_1 \Sigma^{(1)} + a_2 \Sigma^{(2)})^{-1}}= \frac{1}{n_1 + n_2} \bigtr{(a_1 M^{\top} M + a_2 \id)^{-1}}. \]
%As we are going to show later, covariate shift is accurately captured by the spectrum of $\Sigma^{1/2}\Sigma^{-1/2}$.
Hence the variance estimate is determined by the spectrum of $M$.

To illustrate this result, suppose that half of $M$'s singular values are equal to $\lambda > 1$ and the other half are equal to $\lambda^{-1}$.
Hence, when $\lambda = 1$, then there is no covariate shift.
As $\lambda$ increases, the severity of covariate shift increases.
We observe the following dichotomy.
\begin{itemize}
	\item If $n_1 \ge n_2$, then the variance estimate is smallest when there is no covariate shift.
	\item If $n_1 < n_2$, then the variance estimate is largest when there is no covariate shift.
\end{itemize}
\end{example}
We explain why the dichotomy happens. The variance estimate for this example is equal to
\begin{align*}
	\frac{p}{2(n_1 + n_2)} f(\lambda), \text{ where } f(\lambda) = {(\lambda^{-2} a_1 + a_2)^{-1} + (\lambda^2 a_1 + a_2)^{-1}}.
\end{align*}
Using the fact that $a_1 + a_2 = 1 - \frac{p}{n_1 + n_2 - p}$, we have that
\begin{align*}
	f(\lambda) - f(1) &= {(\lambda^2 a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1})^{-1}
	+ (\lambda^{-2} a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1)^{-1}  - \frac{2(n_1 + n_2)}{n_1 + n_2 - p} \\
	&= (2a_1 - \frac{n_1 + n_2} {n_1 + n_2 - p}) g(\lambda, a_1),
\end{align*}
where $g(\lambda, a_1) > 0$ is a fixed function and can be derived from algebraic calculation (details omitted).
%Hence, whether or not $f(\lambda)$ is small than $f(1)$ depends on $a_1$.
We show that $a_1 \ge \frac{n_1 + n_2}{2(n_1 + n_2 - p)}$ if and only if $n_1 \ge n_2$, thus explaining the dichonomy.

In Example \ref{ex_covshift}, $a_1$ and $a_2$ have a symmetric role in the self-consistent equations \eqref{lem_cov_shift_eq}.
Hence, when $n_1 \ge n_2$, we have that $a_1 \ge a_2$, hence $a_1 \ge \frac{1}{2}(1 - \frac{p}{n_1 + n_2 - p}) = \frac{n_1 + n_2}{2 (n_1 + n_2 - p)}$.
The other case when $n_1 < n_2$ is similar.
A formal proof requires solving the self-consistent equations \eqref{lem_cov_shift_eq} and we omit the details.
Thus, we conclude that if $n_1 \ge n_2$, then $f(\lambda) > f(1)$.
If $n_1< n_2$, then $f(\lambda)< f(1)$.
%\begin{proposition}[Source/target sample ratio]\label{prop_data_size}
%	In the isotropic model, suppose that $\rho_1 > 40$ and $\rho_2 > 110$ are fixed constants, and $\Psi(\beta_1, \beta_2) > 2/(\rho_2 - 1)$.
%	Then we have that
%	\begin{itemize}
%		\item \textbf{Positive transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} < \frac{1}{\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1) - \nu^{-1}}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) < \te(\hat{\beta}_2^{\STL}).$$
%		\item \textbf{Negative transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} > {\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 3/2) - \nu}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) > \te(\hat{\beta}_2^{\STL}).$$
%	\end{itemize}
%\end{proposition}
%Proposition \ref{prop_data_size} describes the bias-variance tradeoff in terms of the sample ratio $\rho_1 / \rho_2$.

>>>>

\begin{example}[Covariate shift]\label{ex_covshift}
%So far we have considered the isotropic model where $\Sigma_1 = \Sigma_2$.
%This setting is relevant for settings where different tasks share the same input features such as multi-class image classification.
%In general, the covariance matrices of the two tasks may be different such as in text classification.
Our second example studies the effect of varying covariate shifts on the variance limit in equation \eqref{lem_cov_shift_eq}:
%in the left hand side of equation \eqref{lem_cov_shift_eq}:
	\[ \bigtr{\Sigma^{(2)} \hat{\Sigma}^{-1}} \rightarrow \frac{1}{n_1 + n_2} \bigtr{\Sigma^{(2)} (a_1 \Sigma^{(1)} + a_2 \Sigma^{(2)})^{-1}}= \frac{1}{n_1 + n_2} \bigtr{(a_1 M^{\top} M + a_2 \id)^{-1}}. \]
%As we are going to show later, covariate shift is accurately captured by the spectrum of $\Sigma^{1/2}\Sigma^{-1/2}$.
Hence the variance limit is determined by the spectrum of $M$.

To illustrate this result, suppose that half of $M$'s singular values are equal to $\lambda > 1$ and the other half are equal to $\lambda^{-1}$.
In particular, when $\lambda = 1$, there is no covariate shift.
As $\lambda$ increases, the severity of covariate shift increases.
We observe the following dichotomy.
\begin{itemize}
	\item If $n_1 \ge n_2$, then the variance limit is smallest when there is no covariate shift.
	\item If $n_1 < n_2$, then the variance limit is largest when there is no covariate shift.
\end{itemize}
\end{example}
We explain why the dichotomy happens. The variance estimate for this example is equal to
\begin{align*}
	\frac{p}{2(n_1 + n_2)} f(\lambda), \ \ \text{ where }\ \  f(\lambda) = {(\lambda^{-2} a_1 + a_2)^{-1} + (\lambda^2 a_1 + a_2)^{-1}}.
\end{align*}
Using the fact that $a_1 + a_2 = 1 - \frac{p}{n_1 + n_2 }$, we can obtain that
\begin{align*}
	f(\lambda) - f(1) &= \left(\lambda^2 a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1}
	+ \left(\lambda^{-2} a_1 + \frac{n_1 + n_2 - p}{n_1 + n_2} - a_1\right)^{-1}  - \frac{2(n_1 + n_2)}{n_1 + n_2 - p} \\
	&= \left(2a_1 - \frac{n_1 + n_2-p} {n_1 + n_2 }\right)\cdot (\lambda^2-1)^2  g(\lambda, a_1),
\end{align*}
where $g(\lambda, a_1) > 0$ is a fixed function and can be derived from algebraic calculations (details omitted).
%Hence, whether or not $f(\lambda)$ is small than $f(1)$ depends on $a_1$.
We now show that $a_1 \ge \frac{n_1 + n_2-p}{2(n_1 + n_2 )}$ if and only if $n_1 \ge n_2$, and hence explain the dichonomy. 
%In Example \ref{ex_covshift}, with the fact $M=M^{-1}$, we observe that $a_1$ and $a_2$ play a symmetric role. 
%in the self-consistent equations \eqref{lem_cov_shift_eq}.
In fact, if $a_1>a_2$, then the equations in \eqref{eq_a12extra} give that $a_1> \frac{n_1 + n_2-p}{2 (n_1 + n_2)}$ and
$$ 
 \frac{n_1}{n_1 + n_2}=a_1 + \frac1{n_1 + n_2}\cdot \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2 a_1}{\lambda_i^2 a_1 + a_2}} 
> a_1 + \frac{p}{2(n_1+n_2)} \left(\frac{\lambda^2}{\lambda^2+1}+\frac{\lambda^{-2}}{\lambda^{-2}+1}\right) >\frac{1}{2}. 
$$
This implies $n_1>n_2$. Similarly, if $a_1<a_2$, equations in \eqref{eq_a12extra} give that $a_1 < \frac{n_1 + n_2-p}{2 (n_1 + n_2)}$ and $n_1<n_2$. Thus, we conclude that $f(\lambda) \ge f(1)$ if and only if $n_1 \ge n_2$.

\iffalse
Hence, when $n_1 \ge n_2$, we have that $a_1 \ge a_2$, which gives $a_1 \ge \frac{1}{2}(1 - \frac{p}{n_1 + n_2 }) = \frac{n_1 + n_2-p}{2 (n_1 + n_2)}$.
The other case when $n_1 < n_2$ is similar.
A formal proof follows from a study of the self-consistent equations \eqref{eq_a12extra} and we omit the details.
Thus, we conclude that $f(\lambda) \ge f(1)$ if and only if $n_1 \ge n_2$.  If $n_1< n_2$, then $f(\lambda)< f(1)$.
\fi
%\begin{proposition}[Source/target sample ratio]\label{prop_data_size}
%	In the isotropic model, suppose that $\rho_1 > 40$ and $\rho_2 > 110$ are fixed constants, and $\Psi(\beta_1, \beta_2) > 2/(\rho_2 - 1)$.
%	Then we have that
%	\begin{itemize}
%		\item \textbf{Positive transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} < \frac{1}{\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 1) - \nu^{-1}}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) < \te(\hat{\beta}_2^{\STL}).$$
%		\item \textbf{Negative transfer:} If $\frac{n_1}{n_2} = \frac{\rho_1}{\rho_2} > {\nu} \cdot \frac{1 - 2\rho_2^{-1}}{\Psi(\beta_1, \beta_2) (\rho_2 - 3/2) - \nu}$, then w.h.p. $$\te(\hat{\beta}_2^{\MTL}) > \te(\hat{\beta}_2^{\STL}).$$
%	\end{itemize}
%\end{proposition}
%Proposition \ref{prop_data_size} describes the bias-variance tradeoff in terms of the sample ratio $\rho_1 / \rho_2$.
